Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2019-08-06 scagnostics.Rmd
last of reviewer comments

To find appropriate projections pursuit indexes, literature on variable selection was examined. With high-dimensional data, even plotting all pairs of variables can lead to too many plots, which is what "scagnostics" (@scag, @WW08) were developed to address by providing metrics from which to select the most interesting variable pairs. There are eight scagnostics, of which three ("convex", "skinny" and "stringy") are used here. The question is whether these can be adapted into projection pursuit indexes, to search for unusual features in two-dimensional projections of high-dimensional data. Recent PhD research by @Grimm2016 explored the behaviour of scagnostics for selecting variables, and proposed two more that have nicer properties, based on smoothing splines and distance correlation. In addition, two more indexes for measuring dependence have been proposed in the machine learning literature, based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518], with computationally more efficient versions (MIC_e, TIC_e) [@JMLRv1715308]. These are related to original 1D projection pursuit indexes based on entropy (e.g. @huber85, @JS87). This provides seven current indexes for measuring dependence between two variables, and each is available in an R [@rref] package: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva]. The projection pursuit guided tour is available in the R package, tourr [@tourr], and provides optimisation routines, and visualization.

@@

## Optimization
Given a PPI we are confronted with the task of finding the maximum over all possible $d$ dimensional projections. One challenge is to avoid getting trapped in local maxima that are only a result of sampling fluctuations or a consequence of a noisy index function. @posse95b discusses the optimization, in particular that for most index functions and optimizers results are too local, largely dependent on starting point. @f87 suggested a two-step procedure: the first step is using a large step size to find the approximate global maximum while stepping over pseudomaxima. A second step is then starting from the projection corresponding to the approximate maximum and employing a gradient directed optimization for the identification of the maximum. For exploring high-dimensional data, it can be interesting to observe local maxima as well as a global maximum, and thus a hybrid algorithm that still allows lingering but not being trapped by local maxima is ideal. In addition, being able to visually monitor the optimisation and see the optimal projection in the context of neighbouring projections is useful. This is provided by combining projection pursuit with the grand tour [@CBCH94]. The properties of a suitable optimization algorithm include monotonicity of the index value, a variable step-size to avoid overshooting and to increase the chance of reaching the nearest maximum, and a stopping criterion allowing to move out of a local maximum and into a new search region [@tourr]. A possible algorithm is inspired by simulated annealing and has been described in @lckl2005, this has been implemented in the \texttt{search\_better} and \texttt{search\_better\_random} search functions in the tourr package. The tourr package also provides the \texttt{search\_geodesic} function, which first selects a promising direction by comparing index values between a selected number of small random steps, and then optimises the function over the line along the geodesic in that direction considering projections up to $\pi/4$ away from the current view.

@@

```{r plotV1V2toV3V4, fig.width=6, fig.height=7, fig.cap="PPIs for projections along an interpolation between two nuisance projections. All projections would be nuisance so the PPI are ideally low and smooth, with little difference between sample sizes (solid lines: $n=1000$; dashed: $n=100$). The scagnostic PPIs are noisy. Some indexes have distinct differences in values between sample sizes."}
namesFixed <- str_replace_all(fullRes$name, "sin", "sine")
fullResMelt <- fullRes %>%
  mutate(name = namesFixed) %>%
  mutate(convex = 1-convex) %>%
  rename("1-convex"=convex) %>%
  rowwise() %>%
  mutate(holes = rescaleHoles(holes)) %>%
  gather(PPI, value, -t, -name, -size) %>%
  filter(PPI != "cmass") %>%
  mutate(PPI = fct_relevel(PPI, "holes", "cmass", "1-convex", "skinny", "stringy", "dcor2d", "splines2d", "MIC", "TIC"))

@@

```{r wIntermediate, fig.width=6, fig.height=7, fig.cap="PPIs for projections along an interpolation between nuisance and structured projections, following $x_1$-$x_2$ to $x_1$-$x_5$ to $x_5$-$x_6$ (solid lines: $n=1000$; dashed: $n=100$). The vertical blue line indicates the position of the projection onto $x_1$-$x_5$ in the sequence. Peaks at the end of the sequence indicate the index sees the structure. The scagnostics, MIC and TIC see all three structures, so are more flexible for general pattern detection. Holes only responds to the pipe, and is a multimodal function for this data with a local maximum at $x_1$-$x_5$. "}
intermediatePlane <- which(attributes(interpolate(t3))$new_basis)[2]
namesFixed <- str_replace_all(fullResPlanned$name, "sin", "sine")
fullResMeltPlanned <- fullResPlanned %>%
  mutate(name = namesFixed) %>%
  mutate(convex = 1-convex) %>%
  rename("1-convex"=convex) %>%
  rowwise() %>%
  mutate(holes = rescaleHoles(holes)) %>%
  gather(PPI, value, -t, -name, -size) %>%
  filter(PPI != "cmass") %>%
  mutate(PPI = fct_relevel(PPI, "holes", "1-convex", "skinny", "stringy", "dcor2d", "splines2d", "MIC", "TIC"))
fullResMeltPlanned %>% 
  ggplot(aes(x=t, y=value)) + 
  geom_line(aes(group=size, linetype=factor(size, levels=c("1000", "100")))) + 
  ylim(c(0,1)) +
  geom_vline(xintercept=intermediatePlane, color="blue", alpha=0.3) +
  facet_grid(PPI~name) + 
  theme(legend.position="none") + xlab("Sequence of projections (t)") + ylab("PPI value")
```

## Optimisation check with the guided tour
\label{sec:guided}

@@

The optimization is done in two stages, a scouting phase using method 2, and a refinement stage using method 3.  For the scouting we use $\alpha = 0.5$ and stopping condition of maximum 5000 tries, and we optimised the TIC index.

@@

```{r pipeFirstRun, fig.width=7, fig.height=7, out.width = "0.8\\textwidth", fig.cap="Scouting for the pipe using optimization method 2 on the TIC index, other PPI values also shown for interest. Only the values for the target planes are shown. Despite the small maximum value of TIC for this data, it identifies the pipe."}
if(!file.exists("cache/pipeFirstRun.rda")){
  pipeTourRes <- getProj(pipe1000, pipeTourFull, "Pipe", 1000)
  #pipeTourResMelt <- melt(pipeTourRes, id=c("t", "name", "size"))
  #never used?
  #save(pipeTourRes, pipeTourResMelt, file = "cache/pipeFirstRun.rda")
  save(pipeTourRes, file = "cache/pipeFirstRun.rda")
} else {
  load("cache/pipeFirstRun.rda")
}
newBasisPipeTour <- pipeTourRes %>%
  mutate(convex = 1-convex) %>%
  rename("1-convex"=convex) %>%
  rowwise() %>%
  mutate(holes = rescaleHoles(holes)) %>%
  select(-cmass) %>%
  gather(variable, value, -t, -name, -size) %>%
  filter(t %in% which(attributes(interpolate(pipeTour))$new_basis)) %>%
  mutate(variable = fct_relevel(variable, "holes", "1-convex", "skinny", "stringy", "dcor2d", "splines2d", "MIC", "TIC"))

@@

Given the patterns in Figure \ref{fig:wIntermediate} it would be expected that the sine could be found easily, using only optimization method 3 with the splines2d, dcor2d, MIC or TIC indexes. This is examined in Figure \ref{fig:findsine}. Optimization is conducted using the splines2d index, and the trace of the PPI over the optimisation is shown, along with the PPI values for the other indexes over that path. The vertical blue lines indicate anchor bases, where the optimizer stops, and does a new search. The distance between anchor planes is smaller as the maximum is neared.

@@

```{r findsine, fig.width=6, fig.height=7, out.width = "0.8\\textwidth", fig.cap="Guided tour optimising  the splines2d index, using method 3, for the sine data, with $n=1000$. Anchor planes are marked by the blue vertical lines, and are closer to each other approaching the maxima. The sine is found relatively easily, by splines2d, and it is indicated that MIC, TIC, dcor2d and convex would also likely find this structure.", results="hide"}
set.seed(2018)
if(!file.exists("cache/findsine.rda")){
  sineResc <- as.matrix(sin1000)
  sineTour <- save_history(sineResc,
                          guided_tour(splineIndex()))
  sineTourFull <- as.list(interpolate(sineTour))
  sineTourRes <- getProj(sin1000, sineTourFull, "Sine", 1000)
  #sineTourResMelt <- melt(sineTourRes, id=c("t", "name", "size"))
  #save(sineResc, sineTour, sineTourFull, sineTourRes, sineTourResMelt, file = "cache/findsine.rda")
  save(sineResc, sineTour, sineTourFull, sineTourRes, file = "cache/findsine.rda")
} else {
  load("cache/findsine.rda")
}
anchorIdx <- as_tibble(which(attributes(interpolate(sineTour))$new_basis))
sineTourResMelt <-sineTourRes %>%
  mutate(convex = 1-convex) %>%
  rename("1-convex"=convex) %>%
  rowwise() %>%
  mutate(holes = rescaleHoles(holes)) %>%
  select(-cmass) %>%
  gather(variable, value, -t, -name, -size) %>%
  mutate(variable = fct_relevel(variable, "holes", "1-convex", "skinny", "stringy", "dcor2d", "splines2d", "MIC", "TIC"))

@@

The spiral is the most challenging structure to detect because it has a small squint angle [@posse95a], especially as the ratio of noise to structure dimensions increases. This is explored using optimisation method 2 to scout the space for approximate maxima. The skinny scagnostic index is used because it was observed (Figure \ref{fig:wIntermediate}) to be sensitive to this structure, although the noisiness of the index might be problematic. The stringy appears to be more sensitive to the spiral, but it has a much smaller squint angle.

The search is conducted for $p=4,5,6$ which would correspond to 2, 3 and 4 noise dimensions respectively. In addition we examine the distance between planes, using a Frobenius norm, as defined by Equation 2 of @BCAH05, and available in the ```proj_dist``` function in the tourr package, to compare searches across dimensions. The distance between planes is related to squint angle, how far away from the ideal projection can the structure be glimpsed. We estimate the squint angle depending on the number of noise dimensions in the appendix. In order for the optimizer to find the spiral, the distance between planes would need to be smaller than the squint angle. Figure \ref{fig:findspiral} summarizes the results. When $p=4$ the scouting method effectively finds the spiral. Plot (a) shows the side-by-side boxplots of pairwise distances between planes examined during the optimization, for $p=4,5,6$. These are on average smaller for the lower dimension, and gradually increase as dimension increases. This is an indication of the extra computation needed to brute force find the spiral as noise dimensions increase. Plot (b) shows the distance of the plane in each iteration of the optimisation to the ideal plane, where it can be seen that only when $p=4$ does it converge to the ideal. Its likely that expanding the search space should result in uncovering the spiral in higher dimensions, which however requires tuning of the stopping conditions and long run times.

```{r findspiral, out.width="0.9\\textwidth", fig.cap="Guided tour optimising the skinny index for the Sprial dataset with 1000 datapoints, with p = 4, 5, 6. The left plot shows the distribution of pairwise distances between planes obtained via the guided tour, the right shows the evolution of distance to the ideal plane as the index is being optimised.", results="hide", fig.width=7, fig.height=2.5}

@@

- Scaling and standardization of the data
- Index function selection and if necessary formatting according to the guided tour requirements, i.e. a function taking matrix input and returning the index value
- Calling the guided tour, for exploration this can be done via the `tourr::animate` function, for recording the results `tourr::save_history` should be used, examples are given in the documentation of the `tourr::guided_tour` function
- Things to explore: different indexes, different starting planes, different optimisation options, different ways of reducing dimensionality prior to the guided tour run

@@

```{r nseRemL1, fig.cap="Removing L1 and optimise again of the remaining parameters, where m2 remains removed from the set. Because of parameter correlations we can recover clear description of L2 as a function of the other parameters, despite m2 missing."}
nsM2 <- nsD %>%
  select(-m2, -L1) %>% # remove m2 to avoid finding well known correlation
  rescale()
if(!file.exists("cache/neutronRemL1.rda")){
  set.seed(2018)
  nseTour2 <- save_history(nsM2, guided_tour(splineIndex()))
  save(nseTour2, file = "cache/neutronRemL1.rda")
} else{
  load("cache/neutronRemL1.rda")
}
tL <- as.list(nseTour2)
iLast <- length(tL)
fProj <- tL[[iLast]]
dProj <- as_tibble(nsM2 %*% fProj)
colnames(fProj) <- c("PP1", "PP2")
colnames(dProj) <- c("PP1", "PP2")
fProj <- as_tibble(fProj) %>%
  add_column(label=colnames(nsM2))
p1 <- ggplot(as_tibble(dProj), aes(PP1, PP2))+geom_point(alpha=0.1)+ theme(aspect.ratio=1) + 
  geom_path(data=getCircle(0.5,0.6,0.3,0.18), aes(x=x, y=y), color="grey") + 
  xlim(0, 1.5) + ylim(-0.1, 0.8) +
  geom_segment(data=getAxes(fProj, 0.5,0.6,0.3,0.18), aes(x=x1, xend=x2, y=y1, yend=y2)) +
  geom_text(data=getAxes(fProj, 0.5,0.6,0.3,0.18), aes(x=x2, y=y2, label=label), size=2.5)
p2 <- ggplot(as_tibble(nsM2), aes(m1+R1+1.5*R2, L2)) + geom_point(alpha=0.1)+ theme(aspect.ratio=1)
p3 <- ggplot(nsD, aes(R2/m2, L2))+ geom_point(alpha=0.1)+ theme(aspect.ratio=1)
grid.arrange(p1, p2, p3, ncol=3)
```

@@

Choosing a different subset of variables reveals something new. The subspace of m1, ra, chi\_tot, alpha, distance, dec produces a more refined view of Figure \ref{fig:bbhGuided3} projection f. When alpha contributes in contrast to dec, the relationship between the points is almost perfectly on a curve. This is shown in Figure \ref{fig:constructedExample}. Manually reconstructing the optimal projection (left plot) can be done by differencing the two parameters, in their original units. This highlights the importance of improved optimisation, that would use tiny final steps to polish the view to a finer optimal projection and possibly remove noise induced by small contributions of many variables.

@@

```{r randuEx, results="hide", fig.cap="Best projection obtained by optimising the MIC index on the RANDU data, using different number of bins, defined by $\\delta$. The smaller the value the fewer bins.  Above each plot is written the value of $\\delta$, time required to optimize (seconds) and the MIC index value. The best $\\delta = 0.7$, and the result indicates that this parameter does affect MIC performance."}
if(!file.exists("cache/randuEx.rda")){
  pL <- list()
  i <- 1
  for (alpha in c( 0.6, 0.7, 0.8)){
    set.seed(556677)
    tic.clearlog()
    tic() #start timer
    tP <- save_history(randu, guided_tour(mineIndexAlpha("MIC", alpha)),max_bases = 100000)
    toc(log=TRUE,quiet=TRUE)
    optT <- unlist(tic.log(format=FALSE))["toc.elapsed"]-unlist(tic.log(format=FALSE))["tic.elapsed"]
    imax <- length(as.list(tP))
    fV <- as_tibble(as.matrix(randu) %*% as.list(tP)[[imax]])
    colnames(fV) <- c("PP1", "PP2")
    idxV <- mine(fV$PP1, fV$PP2, alpha = alpha)[["MIC"]]
    pL[[i]] <- ggplot(fV, aes(PP1,PP2)) + geom_point() + theme(aspect.ratio=1) + 
      ggtitle(paste(alpha, toString(format(optT, digits=2)), toString(format(idxV, digits=2)), sep = ", "))
    i <- i+1
  }
  save(pL, file = "cache/randuEx.rda")
} else {
  load("cache/randuEx.rda")
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], ncol=3)
```

@@

```{r testpipe,  fig.height=4, fig.width=8, out.width = "0.8\\textwidth",  fig.cap="Projections returned by TIC optimisation: by the scouting phase (left) and refined by  optimisation method 3 (right), starting from the scouting phase projection."}
iLast <- length(pipeTourFull)
fProj <- pipeTourFull[[iLast]]
dProj <- as_tibble(pipeResc %*% fProj)
p1 <- ggplot(dProj, aes(V1,V2)) + geom_point() + theme(aspect.ratio=1) +
  xlab("PP1") + ylab("PP2")
iLast <- length(pipeTourFull2)
fProj <- pipeTourFull2[[iLast]]
dProj <- as_tibble(pipeResc %*% fProj)
p2 <- ggplot(dProj, aes(V1,V2)) + geom_point() + theme(aspect.ratio=1) +
  xlab("PP1") + ylab("PP2")
grid.arrange(p1, p2, ncol=2)
```

>>>>

To find appropriate projections pursuit indexes, literature on variable selection was examined. With high-dimensional data, even plotting all pairs of variables can lead to too many plots, which is what "scagnostics" (@scag, @WW08) were developed to address by providing metrics from which to select the most interesting variable pairs. There are eight scagnostics, of which three ("convex", "skinny" and "stringy") are used here. The question is whether these can be adapted into projection pursuit indexes, to search for unusual features in two-dimensional projections of high-dimensional data. Recent PhD research by @Grimm2016 explored the behaviour of scagnostics for selecting variables, and proposed two more that have nicer properties, based on smoothing splines and distance correlation. In addition, two more indexes for measuring dependence have been proposed in the machine learning literature, based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518], with computationally more efficient versions (MIC_e, TIC_e) [@JMLRv1715308]. These are related to original 1D projection pursuit indexes based on entropy (e.g. @huber85, @JS87). This provides seven current indexes for measuring dependence between two variables, and each is available in an R [@rref] package: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva]. The projection pursuit guided tour is available in the R package, tourr [@tourr], and provides optimization routines, and visualization.

@@

## Optimization
Given a PPI we are confronted with the task of finding the maximum over all possible $d$ dimensional projections. One challenge is to avoid getting trapped in local maxima that are only a result of sampling fluctuations or a consequence of a noisy index function. @posse95b discusses the optimization, in particular that for most index functions and optimizers results are too local, largely dependent on starting point. @f87 suggested a two-step procedure: the first step is using a large step size to find the approximate global maximum while stepping over pseudomaxima. A second step is then starting from the projection corresponding to the approximate maximum and employing a gradient directed optimization for the identification of the maximum. For exploring high-dimensional data, it can be interesting to observe local maxima as well as a global maximum, and thus a hybrid algorithm that still allows lingering but not being trapped by local maxima is ideal. In addition, being able to visually monitor the optimization and see the optimal projection in the context of neighbouring projections is useful. This is provided by combining projection pursuit with the grand tour [@CBCH94]. The properties of a suitable optimization algorithm include monotonicity of the index value, a variable step-size to avoid overshooting and to increase the chance of reaching the nearest maximum, and a stopping criterion allowing to move out of a local maximum and into a new search region [@tourr]. A possible algorithm is inspired by simulated annealing and has been described in @lckl2005, this has been implemented in the \texttt{search\_better} and \texttt{search\_better\_random} search functions in the tourr package. The tourr package also provides the \texttt{search\_geodesic} function, which first selects a promising direction by comparing index values between a selected number of small random steps, and then optimizes the function over the line along the geodesic in that direction considering projections up to $\pi/4$ away from the current view.

@@

```{r plotV1V2toV3V4, fig.width=6, fig.height=7, fig.cap="PPIs for  projections along a geodesic interpolation between two nuisance projections. All projections would be nuisance so the PPI are ideally low and smooth, with little difference between sample sizes (solid lines: $n=1000$; dashed: $n=100$). The scagnostic PPIs are noisy. Some indexes have distinct differences in values between sample sizes. (This is not an optimization path, but an interpolation containing 41 projections between two known projections.)"}
namesFixed <- str_replace_all(fullRes$name, "sin", "sine")
fullResMelt <- fullRes %>%
  mutate(name = namesFixed) %>%
  mutate(convex = 1-convex) %>%
  rename("1-convex"=convex) %>%
  rowwise() %>%
  mutate(holes = rescaleHoles(holes)) %>%
  gather(PPI, value, -t, -name, -size) %>%
  filter(PPI != "cmass") %>%
  mutate(PPI = fct_relevel(PPI, "holes", "cmass", "1-convex", "skinny", "stringy", "dcor2d", "splines2d", "MIC", "TIC"))

@@

```{r wIntermediate, fig.width=6, fig.height=7, fig.cap="PPIs for projections along an interpolation between nuisance and structured projections, following $x_1$-$x_2$ to $x_1$-$x_5$ to $x_5$-$x_6$ (solid lines: $n=1000$; dashed: $n=100$). The vertical blue line indicates the position of the projection onto $x_1$-$x_5$ in the sequence. Peaks at the end of the sequence indicate the index sees the structure. The scagnostics, MIC and TIC see all three structures, so are more flexible for general pattern detection. Holes only responds to the pipe, and is a multimodal function for this data with a local maximum at $x_1$-$x_5$. (This is not an optimization path, but an interpolation containing 59 projections between three known projections.) "}
intermediatePlane <- which(attributes(interpolate(t3))$new_basis)[2]
namesFixed <- str_replace_all(fullResPlanned$name, "sin", "sine")
fullResMeltPlanned <- fullResPlanned %>%
  mutate(name = namesFixed) %>%
  mutate(convex = 1-convex) %>%
  rename("1-convex"=convex) %>%
  rowwise() %>%
  mutate(holes = rescaleHoles(holes)) %>%
  gather(PPI, value, -t, -name, -size) %>%
  filter(PPI != "cmass") %>%
  mutate(PPI = fct_relevel(PPI, "holes", "1-convex", "skinny", "stringy", "dcor2d", "splines2d", "MIC", "TIC"))
fullResMeltPlanned %>% 
  ggplot(aes(x=t, y=value)) + 
  geom_line(aes(group=size, linetype=factor(size, levels=c("1000", "100")))) + 
  ylim(c(0,1)) +
  geom_vline(xintercept=intermediatePlane, color="blue", alpha=0.3) +
  facet_grid(PPI~name) + 
  theme(legend.position="none") + xlab("Sequence of projections (t)") + ylab("PPI value")
```

## Optimization check with the guided tour
\label{sec:guided}

@@

The optimization is done in two stages, a scouting phase using method 2, and a refinement stage using method 3.  For the scouting we use $\alpha = 0.5$ and stopping condition of maximum 5000 tries, and we optimized the TIC index.

@@

```{r pipeFirstRun, fig.width=7, fig.height=7, out.width = "0.8\\textwidth", fig.cap="PPIs for a sequence of projections produced by scouting for the pipe using optimization method 2 on the TIC index. Other PPI values are shown for interest. Only the values on the target planes are shown. Despite the small maximum value of TIC for this data, it identifies the pipe."}
if(!file.exists("cache/pipeFirstRun.rda")){
  pipeTourRes <- getProj(pipe1000, pipeTourFull, "Pipe", 1000)
  #pipeTourResMelt <- melt(pipeTourRes, id=c("t", "name", "size"))
  #never used?
  #save(pipeTourRes, pipeTourResMelt, file = "cache/pipeFirstRun.rda")
  save(pipeTourRes, file = "cache/pipeFirstRun.rda")
} else {
  load("cache/pipeFirstRun.rda")
}
newBasisPipeTour <- pipeTourRes %>%
  mutate(convex = 1-convex) %>%
  rename("1-convex"=convex) %>%
  rowwise() %>%
  mutate(holes = rescaleHoles(holes)) %>%
  select(-cmass) %>%
  gather(variable, value, -t, -name, -size) %>%
  filter(t %in% which(attributes(interpolate(pipeTour))$new_basis)) %>%
  mutate(variable = fct_relevel(variable, "holes", "1-convex", "skinny", "stringy", "dcor2d", "splines2d", "MIC", "TIC"))

@@

Given the patterns in Figure \ref{fig:wIntermediate} it would be expected that the sine could be found easily, using only optimization method 3 with the splines2d, dcor2d, MIC or TIC indexes. This is examined in Figure \ref{fig:findsine}. Optimization is conducted using the splines2d index, and the trace of the PPI over the optimization is shown, along with the PPI values for the other indexes over that path. The vertical blue lines indicate anchor bases, where the optimizer stops, and does a new search. The distance between anchor planes is smaller as the maximum is neared.

@@

```{r findsine, fig.width=6, fig.height=7, out.width = "0.8\\textwidth", fig.cap="PPIs for sequence of projections produced by a guided tour optimizing  the splines2d index, using optimization method 3, for the sine data, with $n=1000$. Anchor planes are marked by the blue vertical lines, and are closer to each other approaching the maxima. The sine is found relatively easily, by splines2d, and it is indicated that MIC, TIC, dcor2d and convex would also likely find this structure.", results="hide"}
set.seed(2018)
if(!file.exists("cache/findsine.rda")){
  sineResc <- as.matrix(sin1000)
  sineTour <- save_history(sineResc,
                          guided_tour(splineIndex()))
  sineTourFull <- as.list(interpolate(sineTour))
  sineTourRes <- getProj(sin1000, sineTourFull, "Sine", 1000)
  #sineTourResMelt <- melt(sineTourRes, id=c("t", "name", "size"))
  #save(sineResc, sineTour, sineTourFull, sineTourRes, sineTourResMelt, file = "cache/findsine.rda")
  save(sineResc, sineTour, sineTourFull, sineTourRes, file = "cache/findsine.rda")
} else {
  load("cache/findsine.rda")
}
anchorIdx <- as_tibble(which(attributes(interpolate(sineTour))$new_basis))
sineTourResMelt <-sineTourRes %>%
  mutate(convex = 1-convex) %>%
  rename("1-convex"=convex) %>%
  rowwise() %>%
  mutate(holes = rescaleHoles(holes)) %>%
  select(-cmass) %>%
  gather(variable, value, -t, -name, -size) %>%
  mutate(variable = fct_relevel(variable, "holes", "1-convex", "skinny", "stringy", "dcor2d", "splines2d", "MIC", "TIC"))

@@

The spiral is the most challenging structure to detect because it has a small squint angle [@posse95a], especially as the ratio of noise to structure dimensions increases. This is explored using optimization method 2 to scout the space for approximate maxima. The skinny scagnostic index is used because it was observed (Figure \ref{fig:wIntermediate}) to be sensitive to this structure, although the noisiness of the index might be problematic. The stringy appears to be more sensitive to the spiral, but it has a much smaller squint angle.

The search is conducted for $p=4,5,6$ which would correspond to 2, 3 and 4 noise dimensions respectively. In addition we examine the distance between planes, using a Frobenius norm, as defined by Equation 2 of @BCAH05, and available in the ```proj_dist``` function in the tourr package, to compare searches across dimensions. The distance between planes is related to squint angle, how far away from the ideal projection can the structure be glimpsed. We estimate the squint angle depending on the number of noise dimensions in the appendix. In order for the optimizer to find the spiral, the distance between planes would need to be smaller than the squint angle. Figure \ref{fig:findspiral} summarizes the results. When $p=4$ the scouting method effectively finds the spiral. Plot (a) shows the side-by-side boxplots of pairwise distances between planes examined during the optimization, for $p=4,5,6$. These are on average smaller for the lower dimension, and gradually increase as dimension increases. This is an indication of the extra computation needed to brute force find the spiral as noise dimensions increase. Plot (b) shows the distance of the plane in each iteration of the optimization to the ideal plane, where it can be seen that only when $p=4$ does it converge to the ideal. Its likely that expanding the search space should result in uncovering the spiral in higher dimensions, which however requires tuning of the stopping conditions and long run times.

```{r findspiral, out.width="0.9\\textwidth", fig.cap="Guided tour optimizing the skinny index for the Sprial dataset with 1000 datapoints, with p = 4, 5, 6. The left plot shows the distribution of pairwise distances between planes obtained via the guided tour, the right shows the evolution of distance to the ideal plane as the index is being optimized.", results="hide", fig.width=7, fig.height=2.5}

@@

- Scaling and standardization of the data
- Index function selection and if necessary formatting according to the guided tour requirements, i.e. a function taking matrix input and returning the index value
- Calling the guided tour, for exploration this can be done via the `tourr::animate` function, for recording the results `tourr::save_history` should be used, examples are given in the documentation of the `tourr::guided_tour` function
- Things to explore: different indexes, different starting planes, different optimization options, different ways of reducing dimensionality prior to the guided tour run

@@

```{r nseRemL1, fig.cap="Removing L1 and optimize again of the remaining parameters, where m2 remains removed from the set. Because of parameter correlations we can recover clear description of L2 as a function of the other parameters, despite m2 missing."}
nsM2 <- nsD %>%
  select(-m2, -L1) %>% # remove m2 to avoid finding well known correlation
  rescale()
if(!file.exists("cache/neutronRemL1.rda")){
  set.seed(2018)
  nseTour2 <- save_history(nsM2, guided_tour(splineIndex()))
  save(nseTour2, file = "cache/neutronRemL1.rda")
} else{
  load("cache/neutronRemL1.rda")
}
tL <- as.list(nseTour2)
iLast <- length(tL)
fProj <- tL[[iLast]]
dProj <- as_tibble(nsM2 %*% fProj)
colnames(fProj) <- c("PP1", "PP2")
colnames(dProj) <- c("PP1", "PP2")
fProj <- as_tibble(fProj) %>%
  add_column(label=colnames(nsM2))
p1 <- ggplot(as_tibble(dProj), aes(PP1, PP2))+geom_point(alpha=0.1)+ theme(aspect.ratio=1) + 
  geom_path(data=getCircle(0.5,0.6,0.3,0.18), aes(x=x, y=y), color="grey") + 
  xlim(0, 1.5) + ylim(-0.1, 0.8) +
  geom_segment(data=getAxes(fProj, 0.5,0.6,0.3,0.18), aes(x=x1, xend=x2, y=y1, yend=y2)) +
  geom_text(data=getAxes(fProj, 0.5,0.6,0.3,0.18), aes(x=x2, y=y2, label=label), size=2.5)
p2 <- ggplot(as_tibble(nsM2), aes(m1+R1+1.5*R2, L2)) + geom_point(alpha=0.1)+ theme(aspect.ratio=1)
p3 <- ggplot(nsD, aes(R2/m2, L2))+ geom_point(alpha=0.1)+ theme(aspect.ratio=1)
grid.arrange(p1, p2, p3, ncol=3)
```

@@

Choosing a different subset of variables reveals something new. The subspace of m1, ra, chi\_tot, alpha, distance, dec produces a more refined view of Figure \ref{fig:bbhGuided3} projection f. When alpha contributes in contrast to dec, the relationship between the points is almost perfectly on a curve. This is shown in Figure \ref{fig:constructedExample}. Manually reconstructing the optimal projection (left plot) can be done by differencing the two parameters, in their original units. This highlights the importance of improved optimization, that would use tiny final steps to polish the view to a finer optimal projection and possibly remove noise induced by small contributions of many variables.

@@

```{r randuEx, results="hide", fig.cap="Best projection obtained by optimizing the MIC index on the RANDU data, using different number of bins, defined by $\\delta$. The smaller the value the fewer bins.  Above each plot is written the value of $\\delta$, time required to optimize (seconds) and the MIC index value. The best $\\delta = 0.7$, and the result indicates that this parameter does affect MIC performance."}
if(!file.exists("cache/randuEx.rda")){
  pL <- list()
  i <- 1
  for (alpha in c( 0.6, 0.7, 0.8)){
    set.seed(556677)
    tic.clearlog()
    tic() #start timer
    tP <- save_history(randu, guided_tour(mineIndexAlpha("MIC", alpha)),max_bases = 100000)
    toc(log=TRUE,quiet=TRUE)
    optT <- unlist(tic.log(format=FALSE))["toc.elapsed"]-unlist(tic.log(format=FALSE))["tic.elapsed"]
    imax <- length(as.list(tP))
    fV <- as_tibble(as.matrix(randu) %*% as.list(tP)[[imax]])
    colnames(fV) <- c("PP1", "PP2")
    idxV <- mine(fV$PP1, fV$PP2, alpha = alpha)[["MIC"]]
    pL[[i]] <- ggplot(fV, aes(PP1,PP2)) + geom_point() + theme(aspect.ratio=1) + 
      ggtitle(paste(alpha, toString(format(optT, digits=2)), toString(format(idxV, digits=2)), sep = ", "))
    i <- i+1
  }
  save(pL, file = "cache/randuEx.rda")
} else {
  load("cache/randuEx.rda")
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], ncol=3)
```

@@

```{r testpipe,  fig.height=4, fig.width=8, out.width = "0.8\\textwidth",  fig.cap="Projections returned by TIC optimization: by the scouting phase (left) and refined by  optimization method 3 (right), starting from the scouting phase projection."}
iLast <- length(pipeTourFull)
fProj <- pipeTourFull[[iLast]]
dProj <- as_tibble(pipeResc %*% fProj)
p1 <- ggplot(dProj, aes(V1,V2)) + geom_point() + theme(aspect.ratio=1) +
  xlab("PP1") + ylab("PP2")
iLast <- length(pipeTourFull2)
fProj <- pipeTourFull2[[iLast]]
dProj <- as_tibble(pipeResc %*% fProj)
p2 <- ggplot(dProj, aes(V1,V2)) + geom_point() + theme(aspect.ratio=1) +
  xlab("PP1") + ylab("PP2")
grid.arrange(p1, p2, ncol=2)
```
